---
title: "README"
author: "Daniel Duplisea"
date: "17/09/2020"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sealevel)
```

# PSMSL sealevel rate of change

This package queries the PSMSL database  (PSMSL 2020) for data from a station of your choosing, plots the RLR monthly data and determines the
rate of sea level change in mm/year.

# Installation

```{r inst, eval=FALSE}
devtools::install_github("duplisea/sealevel")
library(sealevel)
```

# Using the package

## Find a station id based on fuzzy name search

A simple searching function is provided to find stations. PSMSL contains close to 2000 RLR tide gauge stations globally, this is actually not that many. A simple dot plot is provided to show you where these stations are located. In this example we search
for Yarmouth keeping in mind that most nations that were once British colonies probably have a "Yarmouth" which is probably located
at the mouth of a River Yar; however, in the case, searching "yarm" only finds the Yarmouth in Nova Scotia, Canada. You will see from the output to the console that its station id is 1158.


```{r findstations, echo=FALSE}
find.stations.f("yarm", plot=T)
```
## Plotting the sea level data from a station

So let's plot the sea level and calculate its rate of change for Yarmouth (station id 1158). There are few places that have tide level data <1900 but this old fishing port may actually have some so let's set y1 to 1800 just to be sure. Let's set y2 to 2000 
which will then calculate the rate of sea level rate of change from 2000 until the most recent year that data was reported.

```{r yarmouth, echo=FALSE}
sealevel.f(1158, 1800, 2000)
```

Three plots come from this

Plot 1: the sea level at the Yarmouth tide station (1158) from the start of the data (1965) until 2019. A linear regression is fitted to the data and the slope is the rate of sea level change over that time period. In this case that is 3.7 mm/y

Plot 2: this is sea level rate of change since 2000 which is 5.9 mm/y

Plot 3: This takes the full time series of data like in Plot 1 but it fits a segmented regression and determines the breakpoint
that explains the data the best. This has been constrained to fit only 1 breakpoint. The breakpoint is the vertical solid grey line and its 95% confidence interval is represented by the vertical dotted lines. The rate of sea level before and after the breakpoint
are calculated and show on the graph.

# Interpretation

I use this to look at the net sea level rise at a location. Because these are data from fixed tide gauges they simply measure sea level relative to the land at a location and they do not represent global sea level change. The rate of sea level change at a location is a function of local condition which includes glacial rebound rate of the land, subduction and all those other interesting oceanographic and geologic phenomena.

In the case of Yarmouth, Nova Scotia, Canada you can see that the sea has definitely been rising quickly in the past 50 years such that the 50 year average is near to the global present average sea level rise. If you look just from 2000 onward the sea level in Yarmouth has been rising at 5.9 mm/y. This is so fast for a couple of reasons:

1) The Gulf of Maine is a hotspot for sea level rise (Sallenger et al. 2012)
2) 2009/10 saw an AMOC low which created a large rise in that period in northeastern North America (Ezer 2015)
3) This part of North America is presently sinking owing to resettling after glacial rebound as it was closer to the edge of the Laurentide icesheet

Another interesting phenomenon in the decline in sealevel from 1965-1979 shown by the breakpoint regression. This is now at least partially explicable by the phenomenon of dam building and the creation of huge reservoirs of water during the great dam building period in the 1960s and 1970s (Hawley et al. 2020).


# References
Ezer, T., 2015. Detecting changes in the transport of the Gulf Stream and the Atlantic overturning circulation from coastal sea level data: The extreme decline in 2009–2010 and estimated variations for 1935–2012. Global and Planetary Change, 129, pp.23-36.

Hawley, W.B., Hay, C.C., Mitrovica, J.X. and Kopp, R.E., 2020. A Spatially Variable Time Series of Sea Level Change Due to Artificial Water Impoundment. Earth's Future, p.e2020EF001497.

PSMSL. 2020. Permanent service for mean sea level. https://www.psmsl.org/data/.

Sallenger, A.H., Doran, K.S. and Howd, P.A., 2012. Hotspot of accelerated sea-level rise on the Atlantic coast of North America. Nature Climate Change, 2(12), pp.884-888.
